---
title: "Custom order of factor labels"
author: "Jaynal Abedin"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **Custom order of factor labels**

Working with factors in R sometimes might not intuitive. Particularly, if we convert a character variable into factor then the default sort order follows as per alphabetic order. The problem arises when we want to re-arrange the factor labels in the visualisations. This is straight forward for an ordered factor (ordinal variable), but not for un-ordered (nominal variable). In the example below, I will write a short example showing how we can re-arrange factor labels for a nominal variable.

## Example

The first part of this example is to create a data with two variables, as follows:

-   `personId`: A numeric unique ID of study individuals

-   `sportsName`: Name of sports she/he is wanted to build career. This is a character variable

```{r}
set.seed(12345)
exampleDF <- data.frame(
  personId = 1:150,
  sportsName = sample(
    c("Soccer", "Cricket", "Basketball", "Rugby", "Football"),
    size = 150,
    replace = TRUE
  )
)
```

In the code block above, `set.seed()` function used so that the results are reproducible. The data contains 150 rows with unique ID as the sequential numbers, and the sports names were created by taking a sample of size 150 with replacement from five different sports options `c("Soccer", "Cricket", "Basketball", "Rugby", "Football")`. Now the data is ready and we want to create a factor variable from `sportsName` and will create a bar chart using `ggplot2`. Here I will use `dplyr` library for data processing.

```{r, warning=FALSE, message=FALSE}
library(dplyr)

exampleDF <- exampleDF %>%
  mutate(
    sportsFct = as.factor(sportsName)
  )

library(ggplot2)

ggplot(data = exampleDF)+
  geom_bar(
    aes(
      x=sportsFct,
      y=after_stat(count)
    )
  )
```

We can see that factor labels were organized as per their alphabetic order. We want to change this order to display it as: `Soccer`, `Football`, `Rugby`, `Basketball`, and `Cricket`. To achieve this goal, we will use another library called `forcats`.

```{r, warning=FALSE, message=FALSE}
library(forcats)

exampleDF <- exampleDF %>%
  mutate(
    sportsFct = fct_relevel(
      sportsFct,
      c("Soccer", "Football", "Rugby", "Basketball", "Cricket")
    )
  )

ggplot(data = exampleDF)+
  geom_bar(
    aes(
      x=sportsFct,
      y=after_stat(count)
    )
  )
```

Now if we want `Cricket` should display as the first element, then:

```{r, warning=FALSE, message=FALSE}
exampleDF <- exampleDF %>%
  mutate(
    sportsFct = fct_relevel(
      sportsFct,
      c("Cricket","Soccer", "Football", "Rugby", "Basketball")
    )
  )

ggplot(data = exampleDF)+
  geom_bar(
    aes(
      x=sportsFct,
      y=after_stat(count)
    )
  )
```

Please leave a comment or questions if you have any. Also, if you want more customisation in this plot, feel free to ask.
